---
title: 'Week 5: Poisson Regression'
subtitle: 'What should residual plots look like?'
#subtitle: "<span style = 'font-size: 90%;'>Sections 1.1-1.3</span>"
author: "Statistical Modeling"
date: "Last updated: `r Sys.Date()`"
#institute: '`r icon::fa("twitter")` AimeeSMcCoy <br> `r icon::fa("envelope")` aimeeschwab-mccoy@creighton.edu'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    lib_dir: libs
    nature:
      titleSlideClass: ['left', 'middle', 'inverse']
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{css, include=FALSE}
@media print {
  .has-continuation {
    display: block !important;
  }
}

```

```{r xaringan-setup, include=FALSE}
library(xaringanthemer)
library(xaringanExtra)
style_duo_accent(primary_color = "#005A6F",
                 secondary_color = "#f1fffe",
  header_font_google = google_font("Source Sans Pro"),
  text_font_google = google_font("Source Sans Pro"))

#xaringanExtra::use_logo(
#  image_url = "https://upload.wikimedia.org/wikipedia/en/thumb/f/f2/Creighton_University_seal.svg/1200px-Creighton_University_seal.svg.png"
#)


xaringanExtra::use_tachyons()

xaringanExtra::use_tile_view()

xaringanExtra::use_fit_screen()

xaringanExtra::use_editable(expires = 1)

#xaringanExtra::use_slide_tone()

xaringanExtra::use_panelset()

xaringanExtra::use_extra_styles(hover_code_line = TRUE, mute_unhighlighted_code = FALSE)
#xaringanExtra::use_extra_styles(mute_unhighlighted_code = TRUE)

knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, cache=FALSE)

library(tidyverse)
library(RColorBrewer)
library(patchwork)
library(kableExtra)
```

## Let's find out

Simulate data under two models:

1. Traditional Poisson
2. Zero-inflated Poisson

What do the residual plots look like?

---

## Traditional Poisson

Consider two explanatory variables, $X_1, X_2$. 

$$\eta = \ln(\lambda) = \beta_0 + \beta_1 X_1 + \beta_2 X_2$$

$$\hat{Y} = e^{\eta}$$

Parameters|Simulation 1|Simulation 2
---|---|----|---
$\beta_0$|0|1
$\beta_1$|0.25|0.25
$\beta_2$|0.5|-0.5

We'll let $X_1 \sim Uniform(0, 5)$ and $X_2 \sim Uniform(0, 5)$.

---

## Traditional Poisson: $X_1, X_2$

```{r, warning=FALSE, message=FALSE, eval=FALSE}
library(tidyverse)
library(patchwork)
x1 <- runif(200, min=0, max=5)
x2 <- runif(200, min=0, max=5)

data <- tibble(x1,x2)
p1 <- data %>% ggplot(aes(x=x1)) + geom_histogram(binwidth=0.5)
p2 <- data %>% ggplot(aes(x=x2)) + geom_histogram(binwidth=0.5)

p1+p2
```

---

## Traditional Poisson: $X_1, X_2$

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)
library(patchwork)
x1 <- runif(200, min=0, max=5)
x2 <- runif(200, min=0, max=5)

data <- tibble(x1,x2)
p1 <- data %>% ggplot(aes(x=x1)) + geom_histogram(binwidth=1, fill='#66c2a5', col='black')
p2 <- data %>% ggplot(aes(x=x2)) + geom_histogram(binwidth=1, fill='#66c2a5', col='black')

p1+p2
```

$$\eta = \ln(\lambda) = 0+0.25 X_1 + 0.5 X_2$$

- Minimimum/maximum values for $\eta$?

---

## Traditional Poisson: $X_1, X_2$

```{r, echo=FALSE}
p1+p2
```

- Minimum/maximum values for $\lambda$?

---

## Traditional Poisson: simulation

Simulation 1: $\beta_0=0, \beta_1=0.25, \beta_2=0.5$

```{r, echo=FALSE}
beta0 <- 0
beta1 <- 0.25
beta2 <- 0.5

eta <- beta0 + beta1*x1 + beta2*x2 
y <- rpois(n=200, lambda=exp(eta))

data <- tibble(x1,x2,eta,y)

p1 <- data %>% ggplot(aes(x=x1, y=y)) + geom_point(col="#66c2a5") + geom_smooth(method='lm', se=FALSE)
p2 <- data %>% ggplot(aes(x=x2, y=y)) + geom_point(col="#66c2a5") + geom_smooth(method='lm', se=FALSE)
p1+p2
```

---

## Traditional Poisson: simulation

Simulation 1: $\beta_0=0, \beta_1=0.25, \beta_2=0.5$

```{r, echo=FALSE}
p1 <- data %>% ggplot(aes(x=x1, y=log(y))) + geom_point(col="#66c2a5")  + geom_smooth(method='lm', se=FALSE)
p2 <- data %>% ggplot(aes(x=x2, y=log(y))) + geom_point(col="#66c2a5")  + geom_smooth(method='lm', se=FALSE)
p1+p2
```

---

## Traditional Poisson: model check

Simulation 1: $\beta_0=0, \beta_1=0.25, \beta_2=0.5$

```{r}
model <- glm(y ~ x1 + x2, data=data, family='poisson')
summary(model)
```

---

## Traditional Poisson: residual plots

Simulation 1: $\beta_0=0, \beta_1=0.25, \beta_2=0.5$

```{r, echo=FALSE}
data <- data %>% mutate(fitted = model$fitted.values,
                        residuals = model$residuals)
resid1 <- data %>% ggplot(aes(x=fitted, y=residuals)) + geom_point(col="#66c2a5") + labs(x='Fitted values', y='Residuals')
resid2 <- data %>% ggplot(aes(sample=residuals)) + stat_qq(col="#66c2a5") + stat_qq_line()
resid1 + resid2
```

---

## Traditional Poisson: residual plots

Simulation 1: $\beta_0=0, \beta_1=0.25, \beta_2=0.5$

> _What's different about the default residual plots_?

```{r, fig.width=8, fig.height=8, echo=FALSE}
par(mfrow=c(2,2))
plot(model)
```


---

## Traditional Poisson: residual plots

Simulation 1: $\beta_0=0, \beta_1=0.25, \beta_2=0.5$

```{r, echo=FALSE}
data %>% ggplot(aes(x=fitted, y=y)) + geom_point(col="#66c2a5") + labs(x='Fitted values', y='Observed values')
```

---

## Traditional Poisson: simulation

Simulation 2: $\beta_0=1, \beta_1=0.25, \beta_2=-0.5$

```{r, echo=FALSE}
beta0 <- 1
beta1 <- 0.25
beta2 <- -0.5

eta <- beta0 + beta1*x1 + beta2*x2
y <- rpois(n=200, lambda=exp(eta))

data2 <- tibble(x1,x2,eta,y)

p3 <- data2 %>% ggplot(aes(x=x1, y=y)) + geom_point(col="#fc8d62")
p4 <- data2 %>% ggplot(aes(x=x2, y=y)) + geom_point(col="#fc8d62")
p3+p4
```

---

## Traditional Poisson: simulation

Simulation 2: $\beta_0=1, \beta_1=0.25, \beta_2=-0.5$

```{r, echo=FALSE}
p3 <- data2 %>% ggplot(aes(x=x1, y=log(y))) + geom_point(col="#fc8d62")
p4 <- data2 %>% ggplot(aes(x=x2, y=log(y))) + geom_point(col="#fc8d62")
p3+p4
```

---

## Traditional Poisson: model check

Simulation 2: $\beta_0=1, \beta_1=0.25, \beta_2=-0.5$

```{r}
model2 <- glm(y ~ x1 + x2, data=data2, family='poisson')
summary(model2)
```

---

## Traditional Poisson: residual plots

Simulation 2: $\beta_0=1, \beta_1=0.25, \beta_2=-0.5$

```{r, echo=FALSE}
data2 <- data2 %>% mutate(fitted = model2$fitted.values,
                        residuals = model2$residuals)
resid3 <- data2 %>% ggplot(aes(x=fitted, y=residuals)) + geom_point(col="#fc8d62") + labs(x='Fitted values', y='Residuals')
resid4 <- data2 %>% ggplot(aes(sample=residuals)) + stat_qq(col="#fc8d62") + stat_qq_line()
resid3+resid4
```

---

## Traditional Poisson: residual plots

Simulation 2: $\beta_0=1, \beta_1=0.25, \beta_2=-0.5$

```{r, fig.width=8, fig.height=8, echo=FALSE}
par(mfrow=c(2,2))
plot(model2)
```

---

## Traditional Poisson: residual plots

Simulation 2: $\beta_0=1, \beta_1=0.25, \beta_2=0.5$

```{r, echo=FALSE}
data2 %>% ggplot(aes(x=fitted, y=y)) + geom_point(col="#fc8d62") + labs(x='Fitted values', y='Observed values')
```

---

## Takeaways

1. Residuals in a Poisson model will _not_ be normally distributed

2. Some weird things can happen in the residual plots

3. Depending on the function you use, the scale might change!








